---
title: "508HW1"
author: "Grace Falken"
date: "`r format(Sys.time(), format='%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pander)
library(ggplot2)
```

## Introduction to the Data

The `swiss` dataset presents fertility and economic indicators of `r nrow(swiss)` Swiss provinces in 1888. For each province, we observe fertility, which ranges from `r min(swiss$Fertility)` to `r round(max(swiss$Fertility), 0)`; the percent occupational involvement agriculture; passage rates of an army examination; population percentage with education beyond primary school; percent Catholic; and infant mortality rates. 

First, it's helpful to know the range and distribution of these variables:

```{r data summary, echo=FALSE}
data(swiss)

pander(summary(swiss), stype="rmarkdown", split.tables=120)

```

It's also helpful to observe how these variables are generally related:


```{r pairwise, echo=FALSE, fig.align='center'}
pairs(swiss, pch = 8, col = "blue",
      main = "Pairwise Relationships of Swiss variables")
```

## Relating Agriculture and Education

Let's say we're particularly interested in the relationship between the percent of the population in the agricultural industry. The correlation between these two variables is `r round(cor(swiss$Agriculture, swiss$Examination),3)`, which is relatively high. As a point of comparison, the correlation between how much of the population is Catholic and the portion of the population working in agriculture is `r round(cor(swiss$Agriculture, swiss$Catholic),3)`.

We can also looks at this in more detail: 

```{r ag and ed, echo=FALSE, fig.align='center', message=FALSE}
ggplot(swiss, aes(x=Education,y=Agriculture)) + geom_point(color="orange2") + geom_smooth()
```


And notice that these variables are more tightly related (smaller standard errors) when the population portion with education beyond primary school is below twenty percent.


