---
title: "Class notes 3/31"
author: "Grace Falken"
date: "`r format(Sys.time(), format='%B %d, %Y')`"
output: 
  html_document:
      toc: true
      theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pander)
```

## Intro and syntax

```{r define object}
new.object<-c(1,2,3,4,5)
```


remember how to $\beta_i=\sqrt{l}$


To save results from running a code chunk, set `cache=TRUE` in the `R` chunk options. To embed code in the document, input the statement you want, such as the new object: `r new.object`.

## Data frames

```{r cars data, echo=FALSE}
data(cars)
head(cars, 5)
str(cars) #tells the structure of a dataframe
summary(cars) #tells you different info for different objects (e.g., regression)
hist(cars$speed) #notice that the dollar sign is an extract function


#better histogram:
hist(cars$dist,
     xlab = "Distance (ft)", # X axis label
     main = "Observed stopping distances of cars") # Title
```

## Doing math

```{r assign}
(dist.mean <- mean(cars$dist)) #putting in parens makes this print the answer as well as assign the object
(speed.mean <- mean(cars$speed))
```


```{r plot}
plot(dist ~ speed, data = cars,
     xlab = "Speed (mph)",
     ylab = "Stopping distance (ft)",
     main = "Speeds and stopping distances of cars",
     pch = 16) # Point size
abline(h = dist.mean, col = "firebrick")
abline(v = speed.mean, col = "cornflowerblue")
```

## Swiss data

```{r}
data(swiss)
pairs(swiss, pch = 8, col = "violet",
      main = "Pairwise comparisons of Swiss variables")
```


## Installing packages

Notice that we installed the pander package. This is loaded above and isn't a part of the markdown document.

Also when you install a package you need to use quo

```{r}
pander(summary(swiss), stype="rmarkdown", split.tables=120)
pander(head(swiss, 5), style="rmarkdown", split.tables=120)
```

