---
title: "notes2"
author: "Grace Falken"
date: "4/7/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## Useful shit

```{r tips}

# comments follow pound symbols ... in rmd bodies this creates headers

#you can save an r object on your computer (you can also save functions this way)
#     save(object,file="new_object.RData") 
#     load("new_object.RData")
#these are saved in the working directory

```

`getwd()`
`setwd("/Users/gracefalken/Desktop/PPM Year 1/Q3 Spring/CSSS 508")`
^ Markdown files automatically set where they're saved as the working directory

For journal articles, Breon suggests using rrtools for journal articles and huskydown for UW dissertations.


To run a `.R` script within an `.Rmd` and `.R`, you can use the `source()` function. (E.g., source the cleaning script from an external script).

## Data and Subsetting

```{r gapminder data}

library(gapminder)
str(gapminder) # notice that factor variables are a variable where numeric values are tied to categories (like a likert scale) ("categorical data with an underlying numeric representation")

#notice these are panel data 

```

We want to use `dplyr` to filter the data into subsets

```{r dplyr, message=FALSE}
library(dplyr)

gapminder %>% filter(country=="Algeria") # gives the subsample of all observations for algeria


```

Packages w the same-named functions as in base R temporarily overwrite the base functions
To use those functions, call the package where the function is: `stats::filter`

And thus now we arrive at discussing the pipe. These read left-to-right:

`gapminder$pop %>% mean() %>% log()` gets the mean of the log of the data, analogous to `mean(log(gapminder$pop))`

For our subsampel command, the statement `country=="Algergia"` returns true and false for this logic statement.

### Logical Operators

`==` tests for equivalency.

`!=` not equal to

`>` and fam

`%in%` checks if equal to one of several values

`&` both conditions need to hold

`|` at least one condition holds (OR)

`!` inverts logical condition


```{r multiple conditions}
gapminder %>% filter(country=="Oman" & year > 1980)

#saving a subset as an object
china <- gapminder %>% filter(country=="China")

```


## Plotting with ggplot2

The core idea here is layered grammar of graphics: we can break up elemenets of a plot into pieces and combine them.

```{r plot base r, fig.align='center'}
plot(lifeExp ~ year, 
     data = china, 
     xlab = "Year", 
     ylab = "Life expectancy",
     main = "Life expectancy in China", 
     col = "red", 
     cex.lab = 1.5,
     cex.main= 1.5,
     pch = 16)
```


Versus the `ggplot2` version

```{r china simple, fig.align='center'}
ggplot(data = china, 
       aes(x = year, y = lifeExp)) +
    geom_point()
```

Add layers using `+`. Aesthetics determine how the layers appear. Set aesthetics using arguments inside of layer functions. Aesthetics also determine how data map to appearances.

### Layers

`ggplot()` initializes & specifies input data
`geom_point()` scatterplot
`geom_line()` lines
`ggtitle`, `xlab()`, `ylab()`: layers of labels
`facet_wrap()`: layer creating separate panels stratified by some factor wrapping around
`facet_grid()`: same idea, but can split by two variables along rows and columns (e.g. `facet_grid(gender ~ age_group)`)
`theme_bw()`: replace default gray background with black-and-white


### Aesthetics

`x`, `y`: coordinate values to use
`color`: set color of elements based on some data value
`group`: describe which points are conceptually grouped together for the plot (often used with lines)
`size`: set size of points/lines based on some data value
`alpha`: set transparency based on some data value

Arguments inside the `aes()` (mapping aesthetics) depend on the data you're using. Arguments inside the layers (setting aesthetics) do not.

### Plots babieeeeee

```{r plots for china, fig.align='center'}
ggplot(data = china,
       aes(x = year, y = lifeExp)) +
  geom_point(color="red", size=3) +
  xlab("Year") +
  ylab("Life expectancy") +
  ggtitle("Life expectancy in China") +
  theme_bw(base_size=12) # changes the background color and makes the base text size bigger
```


Let's say we want to instead plot all the countries

```{r all countries, fig.align='center'}

# switch to lines and need to group by country. drop size=3
# also as an aesthetic add color=continent
ggplot(data = gapminder, 
       aes(x = year, y = lifeExp, 
           group = country, 
           color=continent)) +
  geom_line() +
  xlab("Year") + 
  ylab("Life expectancy") +
  ggtitle("Life expectancy over time") +
  theme_bw(base_size=18)

#split these out:
# drop base_size18 to make the x axis labels legible
# legend.position = assigned coords to move it elsewhere ... or instead to drop the legend
#     write "theme(legend.position = "none")"
ggplot(data = gapminder, 
       aes(x = year, y = lifeExp, 
           group = country, 
           color=continent)) +
  geom_line() +
  xlab("Year") + 
  ylab("Life expectancy") +
  ggtitle("Life expectancy over time") +
  theme_bw() +
  facet_wrap(~ continent) +
  theme(legend.position = c(0.8, 0.25))



```


To store a plot, assign it to a name. This will not display the graph. You can then use this object to add more layers.

Syntactal note: you can enter a line break after `<-` assignment operators, `,` commas (for example between options).


```{r store and add layers, fig.align='center'}

lifeExp_by_year <- 
  ggplot(data = gapminder, 
       aes(x = year, y = lifeExp, 
           group = country, 
           color=continent)) +
  geom_line() +
  xlab("Year") + 
  ylab("Life expectancy") +
  ggtitle("Life expectancy over time") +
  theme_bw() +
  facet_wrap(~ continent) +
  theme(legend.position = "none")

lifeExp_by_year + theme(legend.position ="bottom")



```


When adding additional layers, you can change the axis ranges, make them log or sqrt scale.

Text and ticks can be edited most easily in the `theme()` layer of a graph. Note this is different from `theme_bw()` and should always *follow* those more specific themes governing backgrounds.

```{r change legend name}

lifeExp_by_year +
  theme(legend.position = c(0.8, 0.2)) +
  scale_color_manual(
    name = "Which continent are\nwe looking at?", # \n adds a line break
    values = c("Africa" = "seagreen", "Americas" = "turquoise1", 
               "Asia" = "royalblue", "Europe" = "violetred1", "Oceania" = "yellow"))

```


*Scales* are layers that control how the mapped aesthetics appear. You can modify these with a `scale_[aesthetic]_[option]()` layer where `[aesthetic]` is `color`, `shape`, `linetype`, `alpha`, `size`, `fill`, etc. and `[option]` is something like `manual`, `continuous` or `discrete` (depending on nature of the variable).


When you knit an R Markdown file, any plots you make are automatically saved in the "figure" folder in `.png` format. If you want to save another copy (perhaps of a different file type for use in a manuscript), use `ggsave()`:

`ggsave("I_saved_a_file.pdf", plot = lifeExp_by_year,`
       `height = 3, width = 5, units = "in")`

